---
title: "BodyWeight Linear Mixed Model Analysis Analysis"
author: "Michelle Ha"
date: "`r format(Sys.time(), '%d %B %Y)')`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  urlcolor: blue
  html_document:
    'css:': style.css
    figure_caption: true
    highlight: pygments
    number_sections: true
    smart: false
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, echo=FALSE, message=FALSE}
packages <- c("lme4", "nlme", "tidyverse", "lmerTest", "ggplot2", "emmeans")
invisible(lapply(packages, function(package_name) {
  if (!requireNamespace(package_name)) {
    install.packages(package_name)
  }
  library(package_name, character.only = TRUE)
}))

data("BodyWeight")
```

```{r data}
str(BodyWeight)
```

# Data Visualization

```{r line-plot}
BodyWeight |>
  mutate(Rat1 = factor(as.numeric(Rat))) |>
  ggplot(aes(Time, weight, group = Rat1, color = Rat1)) +
  geom_point() +
  facet_wrap(~Rat1) +
  theme_minimal() +
  labs(x = "Time (Days)", y = "Weight (g)")
```

The plot displays the body weight (in grams) of the rats over 64 days at interval of 7 days.

```{r mean-plot}
# Calculate mean weights for each Time and Diet combination
mean_weights <- BodyWeight |>
  summarize(mean_weight = mean(weight), .by = c("Time", "Diet"))

mean_weights |> ggplot(aes(x = Time, y = mean_weight, color = Diet, group = Diet)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Mean Rat Weight Over Time by Diet",
    x = "Time (Days)",
    y = "Mean Weight (g)",
    color = "Diet"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

The figure showing the mean weight of each type of diet over time displays an increasing trend in the body weight of the rats as the days progress.

# Model Fitting

```{r lmm-interaction}
# get weight at baseline
BodyWeight <- BodyWeight |>
  group_by(Rat) |>
  arrange(weight, .by_group = TRUE) |>
  mutate(initial_weight = first(weight)) |>
  ungroup()

# fit LMM with interaction effect
model1 <- lmer(weight ~ Time * Diet + initial_weight + (1 | Rat), BodyWeight)
summary(model1)
anova(model1)
```

The interaction term `Time*Diet` has p-value \< 0.05 therefore it is reasonable to say the type of die significiantly affect the rats body weight and the effect is consistent over time. The initial weight is also a strong indicator for the body weight.

# Model Comparison

```{r lmm-no-interaction}
# fit LMM with no interaction effect
model2 <- lmer(weight ~ Time + Diet + initial_weight + (1 | Rat), BodyWeight)
```

We will fit a simpler model without the interaction term to see if it could potentially performs as well as `model1`. The simpler model still has `Rat` as a random effect and `initial_weight` as fixed effect, except `Time` and `Diet` are now the fixed effects.

```{r compare-models}
# compare models
anova(model1, model2)
```

P-value of \< 0.05 and lower AIC and BIC indicates that the `model2` with the interaction terms provides a significantly better fit relative to the simpler `mode1`.

# Assumption Checking

```{r normality}
# Normality of residuals
# Extract residuals
residuals <- resid(model1)

# Q-Q plot
qqnorm(residuals)
qqline(residuals)
```

The majority of the points fall close to the diagonal line, with some departures at the tails. The overall shape suggests mild non-normality.

```{r homogenity}
# Homogeneity of Variance
# Fitted values
fitted_values <- fitted(model1)

# Residuals vs Fitted plot
plot(fitted_values, residuals)
abline(h = 0, col = "red")
title("Residuals vs Fitted Values")
```

The spread of residuals appears to vary across the range of fitted values. There's noticeably more spread (larger residuals) for fitted values around 500-600 compared to those around 300-400. This suggests some heteroscedasticity.

```{r independence}
# Residuals over time
plot(BodyWeight$Time, residuals)
abline(h = 0, col = "red")
title("Residuals over Time")
```

The residuals appear to be fairly randomly scattered around the zero line across time. There seem to be a few potential outliers, which might warrant further investigation. Overall, the assumption of independence of residuals is generally met.
